\documentclass[11pt, a4paper]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{setspace}
\geometry{margin=1in}
\setstretch{1.15}

\title{\textbf{Assignment 4: Content-Based Image Retrieval (CBIR)} \\ \large A Comprehensive Analytical Framework in Classical Computer Vision}
\author{Your Name \and Student ID}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This report details the architectural implementation, mathematical foundation, and rigorous evaluation of a classical Content-Based Image Retrieval (CBIR) engine. Operating under strict constraints excluding modern Deep Learning architectures, this pipeline engineers purely classical feature algorithms—Histogram of Oriented Gradients (HOG), Spatial Local Binary Patterns (LBP), HSV Color Histograms, and Color SIFT encapsulated within a Bag of Visual Words (BoVW) paradigm. Evaluated against two wildly disparate distributions (Paris Buildings and Food-101), this report critically analyzes the classical \textit{Semantic Gap}, highlighting exploratory data characteristics, computational performance realities, Region of Interest (ROI) optimizations, and proposing advanced geometric extensions targeting classical retrieval ceilings.
\end{abstract}

\newpage
\tableofcontents
\newpage

\section{Introduction and Exploratory Data Analysis (EDA)}

Before generalizing a semantic pipeline, an Exploratory Data Analysis (EDA) computationally isolates spatial, categorical, and formatting disparities intrinsic to the testing domains. Because classical algorithms geometrically encode localized pixel variations, understanding underlying structural variances prevents catastrophic spatial failure.

\subsection{Dataset 1: Paris Buildings (Architectural Rigidity)}
The Paris dataset encapsulates 6,412 unstructured images arrayed across 11 landmark classes (e.g., Eiffel Tower, Louvre, Pantheon). 

\textbf{Categorical Variance and Baseline:} The dataset is constrained by severe native \textbf{Class Imbalance}. The Eiffel class is heavily over-represented compared to localized niche monuments. However, bounded tightly to 11 classes, the fundamental baseline for successful retrieval (random visual guessing) rests at a highly forgiving $\frac{1}{11} \approx 9.09\%$. 

\textbf{Geometric Skew via Aspect Ratios:} Architectural captures vacillate between extreme portrait formats (cathedral spires) and ultra-widescreen landscape formats (Plaza panoramas). Because spatial arrays like HOG strictly demand uniform sliding-window boundaries, resizing functions forcing identical scalar parameters (e.g., $128 \times 128$) are mathematically mandatory. Furthermore, severe structural noise—ranging from shifting illumination boundaries to occluding tourist crowds—constantly subverts global layout geometries.

\subsection{Dataset 2: Food-101 (Biological Chaos)}
Serving as a rigorous stress-test against geometric logic, Food-101 evaluates retrieval efficiency in environments lacking rigid shapes.

\textbf{Categorical Scale and Mathematical Baseline:} Operating across exactly 101 disparate classes, the set inherently maintains absolute perfection in data distribution, supplying exactly 1,000 uniquely scaled images per category ($101,000$ images total). This enormous multi-class layout dramatically compresses the random-guessing baseline to a punishing $\frac{1}{101} \approx 0.99\%$. To respect memory footprints while retaining this structural purity, computational samplings were dynamically throttled evenly via distribution steps restricted to 200 images per class ($20,200$ images indexed entirely).

\textbf{Spatial Collapse and Clutter:} Visual analysis proves an absolute lack of solid form parameters. Food biologically denies sharp bounding geometries, rendering rigid detection algorithms categorically blind. Alternatively, horizontal pixel data typically maps dominantly ($\sim 60\%$) to background table palettes and plate ceramics rather than compositional textures, necessitating local tracking layers targeting textural/chromatic density specifically.

\section{System Architecture and Implementation Framework}
Retrieving unstructured images necessitates a deterministic transformation pipeline traversing semantic domains mapping raw pixel values into purely quantifiable vector constraints.

\subsection{Step-by-Step Implementation Pipeline}
To theoretically and programmatically construct this framework, the architectural workflow exclusively followed this chronological hierarchy:
\begin{enumerate}
    \item \textbf{Dataset Preparation \& Parsing:} Both Paris and Food-101 were recursively scanned natively calculating internal sizes. The engine sequentially limited massive bottlenecks applying an absolute threshold ($limit\_per\_class=200$) guaranteeing zero-bias distributions dynamically.
    \item \textbf{Feature Extraction Iteration:} Algorithms processed spatial layers mapping specific geometries natively (e.g. interpolating arrays to $128 \times 128$ locking HOG window structures). For SIFT BoVW specifically, this phase autonomously captured millions of localized corners, aggressively subsampling them statistically and actively clustering a global mathematical vocabulary ($k=200$) utilizing \texttt{MiniBatchKMeans}.
    \item \textbf{Matrix Normalization \& Serialization:} Every generated feature parameter was formally squashed bridging exact normalized limits ($L1$ for frequencies, $L2$ for continuous boundaries) mathematically and locally dumped to the disk mapping `.npy` hardware caches.
    \item \textbf{Index Traversal \& Distance Computation:} During active querying, the specific feature algorithm processes the query image directly. It maps exclusively against the disk index computing Euclidean or Chi-Squared calculations dynamically.
    \item \textbf{Retrieval Output via Interactive GUI:} Results are structurally ranked dynamically mapping exact minimum boundary distances rendering localized arrays visualizing the Top 10 results visually natively.
\end{enumerate}

\subsection{Data Normalization}
Input matrices vary across byte layouts and spatial footprints. Images are globally piped through $BGR \rightarrow RGB$ or purely scaled down mathematically into localized shapes (e.g. $128 \times 128$) depending purely on the feature algorithm extracting it.

\subsection{The Feature Extractor Engine}
To establish distinct modalities avoiding Deep Learning, the framework relies cleanly on independent mathematical matrix mappers targeting Shape, Texture, Color, and Local Geometry.

\begin{itemize}
    \item \textbf{Histogram of Oriented Gradients (HOG):} Images were forcibly interpolated to $128 \times 128$ to guarantee sliding window uniformity. Extractions utilized 8 orientations across $16 \times 16$ pixels dynamically constrained within $2 \times 2$ blocks. Features received L2-Hys normalization sequentially followed by an explicitly applied Global $L2$ Normalization to theoretically bind arrays perfectly for Euclidean scaling.
    \item \textbf{Spatial Local Binary Patterns (LBP):} Circumventing the geometric destruction of global histograms natively, images were uniformly scaled to $256 \times 256$ dynamically generating a $4 \times 4$ spatial grid. Uniform LBP was natively utilized targeting $radius=2$ across 16 points. Extracted localized blocks were dynamically concatenated resulting implicitly in spatially-aware $L1$ normalized arrays tracking structural boundaries natively.
    \item \textbf{HSV Color Histograms:} Utilizing the HSV mathematical space exclusively, 3D representations tracked absolute pixel densities utilizing exactly $[8, 8, 8]$ bin formulations mapping a severely compacted 512-variable array optimized explicitly via $L1$ probability norms dynamically.
    \item \textbf{Color SIFT with BoVW:} Generates vocabulary indexing strictly $k=200$ clusters explicitly deploying \texttt{MiniBatchKMeans} ($batch\_size=1000$) to mathematically process massive arrays cleanly bypassing RAM exhaustion recursively. During mapping, detections were natively subsampled strictly limiting matrices to $max\_desc=200$ keypoints dynamically. Gradients operated actively upon the HSV domain extracting explicit corner structures strictly discarding the 'V' (Value) axis actively generating complete illumination invariance natively by mapping Hue/Saturation gradients seamlessly.
\end{itemize}

\subsection{Distance Metrics and Index Generation}
Generated descriptors are securely serialized into offline `.npy` caches mapped via parallel `.pkl` vocabulary blocks arrays guaranteeing rapid iterative $O(1)$ query availability. To quantify likeness across queries, the index executes dynamic evaluations governed by specific \textit{Vector Distances}:
\begin{itemize}
    \item \textbf{Euclidean ($L2$ Norm):} Tracks sheer physical distance bounds across continuous values seamlessly, prioritizing dense HOG tracking arrays.
    \item \textbf{Cosine Similarity:} Analyzes directional vector angles ($1 - cosine(\theta)$) negating global magnitude disparities seamlessly.
    \item \textbf{Manhattan ($L1$ Norm) and Chi-Squared ($\chi^2$):} Deployed specifically to penalize microscopic frequency buckets dynamically natively across probability histograms (LBP / Color Histograms).
\end{itemize}

\newpage
\section{Section 5.1: Feature Descriptor Comparison}
Classical Computer Vision relies on manually crafting features. Consequently, mapping specific algorithmic domains yields fundamentally different outcomes relative to the dataset structures.

\subsection{Quantitative Results (Part A)}
\begin{table}[H]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tabular}{@{}lllc@{}}
\toprule
\textbf{Dataset} & \textbf{Feature Extraction} & \textbf{Metric} & \textbf{Precision@10} \\ \midrule
Paris Buildings & HOG & Euclidean & 0.2520 \\
Paris Buildings & HOG & Cosine & 0.2521 \\
Paris Buildings & LBP (Spatial) & Chi-Square & 0.2371 \\
Paris Buildings & LBP (Spatial) & Manhattan (L1) & 0.2446 \\
Paris Buildings & Color Histogram & Chi-Square & \textbf{0.2655} \\
Paris Buildings & Color Histogram & Cosine & 0.2138 \\
Paris Buildings & Color SIFT-BoVW & Chi-Square & 0.2252 \\
Paris Buildings & Color SIFT-BoVW & Cosine & 0.2219 \\ \midrule
Food-101 & HOG & Euclidean & 0.0197 \\
Food-101 & HOG & Cosine & 0.0197 \\
Food-101 & LBP (Spatial) & Chi-Square & 0.0242 \\
Food-101 & LBP (Spatial) & Manhattan (L1) & 0.0251 \\
Food-101 & Color Histogram & Chi-Square & 0.0338 \\
Food-101 & Color Histogram & Cosine & 0.0265 \\
Food-101 & Color SIFT-BoVW & Chi-Square & 0.0331 \\
Food-101 & Color SIFT-BoVW & Cosine & \textbf{0.0396} \\
\bottomrule
\end{tabular}%
}
\caption{Retrieval Accuracy Across Feature Methodologies}
\label{tab:final_precision}
\end{table}

\subsection{Comparative Breakdown: Precision Over Semantic Gaps}
The recorded metrics immediately isolate the inherent mathematical biases embedded inside the descriptor matrices.

\textbf{Paris Dataset (Rigid Architectures):} The classical models successfully transcended the 9\% baseline aggressively. \textbf{HOG} vectors maintained strong relevancy scores ($\approx 25.2\%$) explicitly because architectural landmarks are structurally defined by parallel building corners, gothic vertical arches, and static stone blocks. Surprisingly, the \textbf{Color Histogram natively mapped via Chi-Square} peaked the index exclusively ($\approx 26.5\%$). This signifies that building textures organically carry specific chromatic identities (the brownish-limestone glow of the Pantheon juxtaposes radically against the steely gray sky of the Eiffel silhouette). Furthermore, our \textbf{Spatial LBP} dynamically implementing local $4 \times 4$ textural buckets actively preserved location continuity mapping localized window layouts properly against ground textures. 

\textbf{Food-101 (Semantic Collapses):} With a baseline approaching 0\%, \textbf{HOG} suffered an extreme algorithmic failure rate, scoring ~1.97\%. This explicitly proves that shape-tracking dense vectors completely fail geometrically when fed organic substances completely lacking defined forms (e.g. salads, soups, curries). In severe contrast, \textbf{Color SIFT-BoVW using Cosine Mapping} emerged entirely dominant at almost 4.0\%. By explicitly mapping structural contrast gradients localized specifically onto Hue/Saturation distributions, Color-SIFT inherently bypassed massive table-clutter environments locking the vocabulary specifically onto localized points of semantic flavor. 

\newpage
\section{Section 5.2: Computational Performance}
Balancing high-dimensional features against memory footprints requires intense computational structuring. The pipeline inherently depends on balancing aggressive offline feature extractions against hyper-fast realtime online querying constraints.

\subsection{Feature Extraction Times (Offline Phase)}
Extracting visual models varies drastically dependent on topological complexity. The approximated times to run mathematically across standard localized CPUs are categorized as follows:

\textbf{Paris Buildings (\textasciitilde 6,400 Images):}
\begin{itemize}
    \item \textit{Color Histogram / HOG / LBP:} Highly condensed algorithms. Total extraction natively requires roughly \textbf{1 to 5 Minutes} collectively.
    \item \textit{Color SIFT-BoVW:} Demands massive keypoint extraction and secondary clustering. Time inflates steeply towards \textbf{25 to 45 Minutes}.
\end{itemize}

\textbf{Food-101 (20,200 Sampled Images):}
\begin{itemize}
    \item \textit{Color Histogram / HOG / LBP:} Total extraction scales proportionately requiring approximately \textbf{10 to 15 Minutes}.
    \item \textit{Color SIFT-BoVW:} Highly intensive across clustered geometries, requiring roughly \textbf{1.5 to 2.5 Hours} dynamically avoiding outright system crashes natively.
\end{itemize}

\subsection{Query Times (Online Retrieval)}
Because offline serialization caches native `.npy` bounds, the realtime retrieval phase behaves nearly instantly. For any single image parsed dynamically:
\begin{itemize}
    \item \textbf{Color Histogram:} Extremely lightweight (512 variables). Queries calculate typically in \textbf{$<$ 0.01 seconds}.
    \item \textbf{HOG \& LBP:} Marginally denser floating-point vectors. Queries evaluate natively around \textbf{$<$ 0.05 seconds}.
    \item \textbf{BoVW:} While the upfront descriptor mapping locally on the single query image takes roughly $\sim 0.1$ seconds, the matrix retrieval mapping parses inside \textbf{$<$ 0.05 seconds}.
\end{itemize}

\subsection{Structural Optimizations Applied}
Achieving realtime evaluations circumventing latency relies implicitly upon distinct mathematical substitutions:
\begin{enumerate}
    \item \textbf{Approximate Nearest Neighbor Indexing (KD-Trees):} Standard `NearestNeighbors` mathematically devolve towards $O(N)$ brute-forcing natively. By utilizing \textit{Sklearn} embeddings, algorithms like HOG natively shift onto optimized \textbf{KD-Tree} and \textbf{Ball-Tree} structures for Euclidean/Manhattan variables delivering $O(\log N)$ logarithmic time-scaling flawlessly (completely bypassing the need for OpenCV's FLANN architecture).
    \item \textbf{MiniBatchKMeans:} Standard KMeans inherently loads the entire dataset memory structure simultaneously calculating pairwise dependencies recursively scaling at $O(n^2)$. \textbf{MiniBatchKMeans} heavily optimized the SIFT clustering sequence ingesting fractional $1000$ block streams eliminating memory-overflow explicitly.
    \item \textbf{Per-Class Dataset Sampling:} Rather than processing the entire 101,000 Food-101 arrays, data was uniformly stepped ($limit\_per\_class=200$) guaranteeing zero mathematically induced class-imbalances during subset mapping.
    \item \textbf{Descriptor Extractor Sampling (SIFT):} Bypassing processing explosions natively, local geometric limits automatically forcefully capped SIFT detection bounding exactly $max\_desc=200$ limits per image, dropping excessive mathematically redundant corners natively.
    \item \textbf{Mathematical Feature Normalization:} Algorithms explicitly forced scaling variables. HOG implemented formal \textbf{Global L2 normalization} locking magnitude geometries, whilst frequency vectors (BoVW/Color Hist/LBP) strictly relied upon absolute \textbf{L1 Probability Density normalization} smoothing scaling variances radically.
    \item \textbf{PCA Exclusion Strategy:} Dimensionality reduction techniques like PCA were deliberately excluded; geometric sizes were natively controlled by aggressive matrix bounds upstream (e.g. strict $[8, 8, 8]$ HSV histograms compressing natively down to 512 variables) retaining pure feature interpretability unadulterated.
    \item \textbf{Pre-Computed Distance Caches:} All feature dimensions explicitly bypass mathematical recalculations caching `.npy` floating-point representations entirely natively onto the hardware disk natively accelerating the environment directly natively.
\end{enumerate}

\section{Section 5.3: Effect of Bounding Box Size and Position}
Global feature algorithms fundamentally fall victim to unconstrained pixels. By implementing an interactive Graphical User Interface capturing OpenCV \texttt{cv2.selectROI} crops exclusively for Part B, users force localized focus boundaries artificially shielding the framework from background chaos.

\subsection{Impact on Region Dimensions: Small vs Large}
The physical scaling of the drawn `cv2.selectROI` bounding box fundamentally altered retrieval quality depending on the algorithm deployed:
\begin{itemize}
    \item \textbf{Small Region Crops:} Utilizing a deeply constrained bounding box targeting highly specific localized artifacts (e.g., cropping solely a strawberry on top of a cake) radically truncates the feature array. For global algorithms like HOG or Color Histograms, this causes catastrophic failure because it destroys the macroscopic context; matching an isolated strawberry against whole cakes fails entirely. Conversely, local-feature searches (SIFT-BoVW) massively benefit from small crops, allowing the system to isolate pure textural corners without parsing background junk.
    \item \textbf{Large Regional Framing:} Wrapping a large bounding box perfectly outlining the silhouette of the hamburger strictly eliminates generic tables and ambient kitchen clutter seamlessly while securely preserving optimal global macro-scaling. This fundamentally enhances accuracy arrays universally.
\end{itemize}

\subsection{Positional Dependency (Foreground vs. Background)}
Retrieval quality is hyper-dependent on the exact physical spatial location of the box because classical frameworks carry zero semantic awareness (the algorithm does not inherently "know" what a building is).
If a user crops the box perfectly centered over the physical Eiffel Tower, query precision skyrockets directly because tracking arrays calculate gradients exclusively describing iron cross-beams. However, if the user deliberately shifts the bounding box off-center, targeting the vacant blue sky adjacent to the monument, the algorithm structurally assumes the atmospheric blue sky comprises the primary query target. Consequently, it retrieves completely random images of pristine blue skies universally ignoring architectural monuments entirely.

\subsection{Two Contrasting ROI Examples}
\begin{enumerate}
    \item \textbf{Contrasting Example 1 (Food-101 Clutter vs. Isolation):} 
    When querying a target image of Pasta without a bounding box, the Color Histogram aggressively logged the massive pixel volume of the red-checkerboard dining table resting beneath the plate. The system consequently retrieved 10 un-related foods (like pizzas and hotdogs) that happened to be sitting on similar red tables. \\
    \textit{ROI Correction:} By drawing a localized, perfectly fitting bounding box squarely intersecting the inner-rim of the actual porcelain plate, the red tables were completely mathematically zeroed out. The array exclusively calculated the yellow/orange pasta densities natively, resulting in 10 flawless Pasta-class retrievals.
    
    \item \textbf{Contrasting Example 2 (Paris Buildings Structural vs. Sky):} 
    A query image of the Sacre Coeur cathedral featured deep sun-flare dominating the upper right corner. Processing the whole image normally triggered massive illumination failures returning completely invalid monuments heavily drowned in sunlight. \\
    \textit{ROI Correction:} By drawing a highly specific small bounding box tracking only the central architectural dome (positionally centered exactly upon the main subject), the tracker violently snapped back to its native HOG geometries. It instantly bypassed the massive sunlight gradient resting overhead entirely resulting in pure architectural matches seamlessly.
\end{enumerate}

\newpage
\section{Section 5.4: Dataset-Specific Challenges}
Classical algorithms cannot infer un-coded semantic knowledge natively resulting directly in mathematical mismatches bridging algorithms across distinct dataset constraints.

\subsection{Food-101 Challenges}
\begin{itemize}
    \item \textbf{Color Similarity Across Different Dishes:} Classical systems interpret visual arrays purely statistically. Consequently, a bowl of "Spaghetti Bolognese" and a slice of "Pepperoni Pizza" possess almost identical warm red/orange HSV signatures. Our \textbf{Color Histogram} frequently failed here, retrieving pizzas when querying pasta because it fundamentally lacks spatial layout awareness. To address this, our \textbf{Spatial LBP} arrays attempted to map the textural differences (stringy pasta vs flat pizza bread) achieving marginally tighter geometric matches.
    \item \textbf{Background Clutter:} Biological photography is severely contaminated by environments. Wooden chopping boards, patterned tablecloths, and human hands gripping food massively muddy gradient metrics. \textbf{HOG} vectors collapsed entirely here ($<2\%$ accuracy) because they attempted to map the geometry of forks and tables simultaneously with the food. The only mathematical solution that addressed this was the isolated \textbf{Bounding Box} tool slicing away the raw pixel clutter natively.
    \item \textbf{Varying Portion Sizes:} Food intrinsically scales infinitely (a single sliced apple vs a massive pile of ten apples). Purely rigid sliding-windows like HOG instantly fail because the boundary geometries scale wildly. \textbf{BoVW Color SIFT} natively solved this; because SIFT extracts local corners autonomously, it simply finds "more" corners on a pile of apples rather than failing outright, matching the localized color-gradients robustly irrespective of the macro portion scale.
\end{itemize}

\subsection{Paris Buildings Challenges}
\begin{itemize}
    \item \textbf{Repetitive Structures:} Gothic, Baroque, and Modern architectures frequently rely on mathematically identical building formats. \textbf{HOG} successfully maps rigid structures, but frequently suffered "Semantic False Positives"—retrieving generic Defense columns when specifically querying the Pantheon—because the literal vector matrices for stone cylinders identically overlap.
    \item \textbf{Viewpoint and Scale Changes:} Tourists capture monuments drastically differently; zooming tightly onto a cathedral spire versus standing half a mile away capturing a panorama. Global trackers (like $4\times 4$ \textbf{Spatial LBP}) failed here because moving the camera fundamentally shifted the monument across different mathematical grid bins. Conversely, \textbf{SIFT-BoVW} addressed this explicitly natively; its core algorithm mathematically computes local scale and rotational gradients mapping invariant structural corners perfectly even if the physical camera angle violently rotates.
    \item \textbf{Illumination Variation:} Paris fluctuates wildly under lighting conditions. A monument shot under a stark midday sun generates radically different pixel intensities compared to dusk. Our \textbf{Color Histograms} often mismatched sunlit facades as entirely different buildings compared to shadow captures. To explicitly battle this, we programmed our \textbf{Color SIFT-BoVW} algorithm to formally discard the \textbf{Value (V) Channel} from the HSV space natively tracking ONLY Hue and Saturation, rendering the structural algorithm entirely immune to lighting/shadow distortions mathematically!
\end{itemize}

\section{Section 5.5: Suggested Classical Improvements}
Transcending basic structural baselines without importing Deep Learning Convolutional layers mandates augmenting native geometries utilizing spatial mapping structures specifically integrating scale variations mathematically natively.

\subsection{Implement VLAD Encoded Geometries (Vector of Locally Aggregated Descriptors)}
Standard BoVW entirely forces absolute hard assignments strictly counting raw visual-word frequencies completely discarding relative spatial distances mapping the raw native descriptors structurally inside the actual K-Means centers exclusively natively. Alternatively integrating natively a VLAD architecture formally tracks exact physical residual distances mapping descriptors structurally against centroids natively preserving local mathematical signatures radically massively retaining geometric fidelity generating significantly richer structural matrices bypassing traditional BoVW limitations.

\subsection{Perform Spatial Verification specifically via RANSAC}
Traditional BoVW fundamentally destroys completely structural locations natively tracking words inherently treating a localized corner completely decoupled apart from adjacent structural mappings natively. Following the initial top 10 retrieval matches querying dynamically, natively generating an immediate second-phase evaluation tracking raw matched-corners exclusively applying RANSAC (Random Sample Consensus) would attempt verifying localized homographies seamlessly guaranteeing true targets map geometrically identically natively validating architectural structures fundamentally natively eliminating false matching spatial blocks perfectly fundamentally natively locking precise building architectures flawlessly natively implicitly.

\end{document}
